IFA Comparison of Noncognitive Skills \\ BCS and MCS

### 2018-05-03 codes/R/paper/cohorts_nc.Rnw
add formula chunk, add 90-10 dec for EXT/INT

read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_scores.R"))
read_chunk(here("codes/R/4_scores_aux.R"))
read_chunk(here("codes/R/5_graphs.R"))
read_chunk(here("codes/R/6_ineq.R"))
read_chunk(here("codes/R/7_counterf.R"))
read_chunk(here("codes/R/7_counterf_graphs.R"))
read_chunk(here("codes/R/7_counterf_tabs.R"))
read_chunk(here("codes/R/7_counterf_seqdec.R"))

@@

<<FA_INV_MAIN, dependson="LOAD_DATA">>=
# main models only
@

% % <<FA_INV_AUX, dependson='LOAD_DATA'>>=
% % # robustness models
% % @

<<FA_FIT, dependson='FA_INV_MAIN'>>=
# assemble fit indices into matrices
@

<<FA_SCORES_MAIN, dependson='FA_INV_MAIN'>>=
# produce scores for main models
@

% % <<FA_SCORES_AUX, dependson='FA_INV_AUX'>>=
% % # produce scores for robustness models
% % @

@@

\clearpage
\begin{figure}
<<FACDENS, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Density of Scored Factors}
{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
\end{figure}

@@

\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data. Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach). In \autoref{fig:meanci}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
%In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile. In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.

\clearpage
\begin{figure}
\begin{center}
<<INCINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=4, fig.width=4, out.width='.5\\linewidth'>>=
@
\end{center}
\caption{Evolution of income inequality }
{\scriptsize \emph{Notes}:  \par}
\end{figure}

\begin{figure}
<<FACINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Mean scores across income quintiles}\label{fig:meanci}
{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
\end{figure}

@@

<<DEC_VARS>>=
# define variables for decomposition
@
% 
% !!!!!!!!!! THIS BREAKS CACHING !!!!!!!!!!!!!!!!!!!!!
% <<DEC_IQR, dependson="DEC_VARS">>=
% # decompose SD and IQRs
% @
% 
% Compare inequality statistics of original vs residualised scores
% <<DEC_RES, dependson='DEC_IQR', echo=T, results='markup'>>=
% iqs.orig
% iqs.resid
% @

\begin{table}
\caption{Quantifying the gradient -- males}
\centering
<<FA_REGS_REAL_M, dependson='DEC_VARS', results='asis'>>=
# regress scores on income
library(dplyr)
library(stargazer)

r.ext.m.bcs <- lm(form.ext, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
r.ext.m.mcs <- lm(form.ext, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
r.int.m.bcs <- lm(form.int, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
r.int.m.mcs <- lm(form.int, data=subset(scores2plot, sex=="M" & cohort=="MCS"))

stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, #r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
          column.labels=c("BCS M","MCS M","BCS M","MCS M"), #,"BCS F","MCS F","BCS F","MCS F"),
          dep.var.caption = "", multicolumn = FALSE,
          omit=c("Constant", paste0("region", c(2:9))),
          # covariate.labels=c(
          #                    "Income Q2",
          #                    "Income Q3",
          #                    "Income Q4",
          #                    "Income Q5"
          #                    ),
          no.space=TRUE,
          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
          notes = "All estimates control for Government Office Region of birth.",
          float=FALSE    # needs begin table outside it
)
@
\end{table}

\begin{table}
\caption{Quantifying the gradient -- females}
\centering
<<FA_REGS_REAL_F, dependson='DEC_VARS', results='asis'>>=
# regress scores on income
library(dplyr)
library(stargazer)

r.ext.m.bcs <- lm(form.ext, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
r.ext.m.mcs <- lm(form.ext, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
r.int.m.bcs <- lm(form.int, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
r.int.m.mcs <- lm(form.int, data=subset(scores2plot, sex=="F" & cohort=="MCS"))

stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, #r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
          column.labels=c("BCS F","MCS F","BCS F","MCS F"), #,"BCS F","MCS F","BCS F","MCS F"),
          dep.var.caption = "", multicolumn = FALSE,
          omit=c("Constant", paste0("region", c(2:9))),
          # covariate.labels=c(
          #                    "Income Q2",
          #                    "Income Q3",
          #                    "Income Q4",
          #                    "Income Q5"
          #                    ),
          no.space=TRUE,
          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
          notes = "All estimates control for Government Office Region of birth.",
          float=FALSE    # needs begin table outside it
)
@
\end{table}

@@

<<COUNTERF, dependson="FA_SCORES_MAIN">>=
# perform counterfactuals
@

<<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
# prepare counterfactual plots
@

\begin{landscape}
\begin{figure}
<<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
cfp.ext
@
\caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf_ext}
{\scriptsize \emph{Notes}: \par}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
<<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
cfp.int
@
\caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf_int}
{\scriptsize \emph{Notes}: \par}
\end{figure}
\end{landscape}

<<COUNTERF_PREP_TAB, dependson='COUNTERF'>>=
# prepare counterfactual tables
@

<<COUNTERF_INEQ, dependson='COUNTERF_PREP_TABS', results='markup'>>=
cat("Males EXT")
print(iqdecomp(logitcf.ext.m))

cat("Males INT")
print(iqdecomp(logitcf.int.m))

cat("Females EXT")
print(iqdecomp(logitcf.ext.f))

cat("Females INT")
print(iqdecomp(logitcf.int.f))
@

@@

\begin{table}[ht]
\caption{Estimated loadings and thresholds} \label{tab:allpars}
\resizebox{\textwidth}{!}{
<<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
source("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/getpars.R")

@@

\begin{table}[ht]
\caption{Estimated latent variable means and covariances} \label{tab:lvpars}
\centering
<<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
require(xtable)
source("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/getpars.R")

@@

>>>>

read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_scores.R"))
read_chunk(here("codes/R/4_scores_aux.R"))
read_chunk(here("codes/R/5_graphs.R"))
read_chunk(here("codes/R/6_decvars.R"))
read_chunk(here("codes/R/7_counterf.R"))
read_chunk(here("codes/R/7_counterf_graphs.R"))
read_chunk(here("codes/R/7_counterf_tabs.R"))
read_chunk(here("codes/R/7_counterf_seqdec.R"))
read_chunk(here("codes/R/getpars.R"))

@@

<<FA_INV_MAIN>>=
# main models only
@

<<FA_FIT>>=
# assemble fit indices into matrices
@

<<FA_SCORES_MAIN>>=
# produce scores for main models
@

@@

\clearpage
\begin{figure}
<<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Density of Scored Factors}
{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
\end{figure}

@@

\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data. Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach). In \autoref{fig:meanci}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.

<<INEQ_STATS, dependson="FA_SCORES_MAIN", results="markup">>=

# income 
tb <- data.frame(
  aggregate(scores2plot$faminc10_infl, by=list(scores2plot$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
)
tb$d9010 <- tb$x[,4]/tb$x[,1]
tb$d7525 <- tb$x[,3]/tb$x[,2]
cat("Quantile differences in income by cohort")
tb

tb.ext <- data.frame(
  aggregate(scores2plot$EXT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
)
tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
tb.ext$d9010 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
cat("\n Quantile differences in EXT by cohort/gender")
tb.ext

tb.int <- data.frame(
  aggregate(scores2plot$INT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
)
tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
tb.int$d9010 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
cat("\n Quantile differences in INT by cohort/gender")
tb.int

\clearpage
\begin{figure}
\begin{center}
<<INCINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=4, fig.width=4, out.width='.5\\linewidth'>>=
@
\end{center}
\caption{Evolution of income inequality }
{\scriptsize \emph{Notes}:  \par}
\end{figure}
% 
% \begin{figure}
% <<FACINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
% @
% \caption{Mean scores across income quintiles}\label{fig:meanci}
% {\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
% \end{figure}

@@

<<FORMULAS>>=
# define variables and formulas for regression and decomposition
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
% \begin{table}
% \caption{Quantifying the gradient}
% \centering
% \resizebox{\textwidth}{!}{
% <<FA_REGS_REAL_M, results='asis'>>=
% # regress scores on income
% library(stargazer)
% 
% decvars_reg <- c(decvars, "cohort") # add cohort
% interact_reg <- paste0(decvars_reg[! decvars_reg %in% c("cohort")], "*cohort")
% 
% form_ext_reg <- as.formula(paste("EXT ~ ", paste0(decvars2, collapse = '+')))
% form.ext2i <- as.formula(paste("EXT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
% form.int2 <- as.formula(paste("INT ~ ", paste0(decvars2, collapse = '+')))
% form.int2i <- as.formula(paste("INT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
% 
% r.ext.m  <- lm(form.ext2,  data=subset(scores2plot, sex=="M"))
% r.exti.m <- lm(form.ext2i, data=subset(scores2plot, sex=="M"))
% r.ext.f  <- lm(form.ext2,  data=subset(scores2plot, sex=="F"))
% r.exti.f <- lm(form.ext2i, data=subset(scores2plot, sex=="F"))
% r.int.m  <- lm(form.int2,  data=subset(scores2plot, sex=="M"))
% r.inti.m <- lm(form.int2i, data=subset(scores2plot, sex=="M"))
% r.int.f  <- lm(form.int2,  data=subset(scores2plot, sex=="F"))
% r.inti.f <- lm(form.int2i, data=subset(scores2plot, sex=="F"))
% 
% stargazer(r.ext.m, r.exti.m, r.ext.f, r.exti.f, r.int.m, r.inti.m, r.int.f, r.inti.f,
%           column.labels=c("M","M","F","F","M","M","F","F"), #,"BCS F","MCS F","BCS F","MCS F"),
%           dep.var.caption = "", multicolumn = FALSE,
%           #omit=c("Constant", paste0("region", c(2:9))),
%           no.space=TRUE,
%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
%           notes = "All estimates control for Government Office Region of birth.",
%           single.row=TRUE,
%           float=FALSE    # needs begin table outside it
% )
% @
% }
% \end{table}

@@

% 
% <<COUNTERF, dependson="FA_SCORES_MAIN">>=
% # perform counterfactuals
% @
% 
% <<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
% # prepare counterfactual plots
% @
% 
% \begin{landscape}
% \begin{figure}
% <<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
% cfp.ext
% @
% \caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf_ext}
% {\scriptsize \emph{Notes}: \par}
% \end{figure}
% \end{landscape}
% 
% \begin{landscape}
% \begin{figure}
% <<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
% cfp.int
% @
% \caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf_int}
% {\scriptsize \emph{Notes}: \par}
% \end{figure}
% \end{landscape}
% 
% <<COUNTERF_PREP_TAB, dependson='COUNTERF'>>=
% # prepare counterfactual tables
% @
% 
% <<COUNTERF_INEQ, dependson='COUNTERF_PREP_TABS', results='markup'>>=
% cat("Males EXT")
% print(iqdecomp(logitcf.ext.m))
% 
% cat("Males INT")
% print(iqdecomp(logitcf.int.m))
% 
% cat("Females EXT")
% print(iqdecomp(logitcf.ext.f))
% 
% cat("Females INT")
% print(iqdecomp(logitcf.int.f))
% @

@@

\begin{table}[ht]
\caption{Estimated loadings and thresholds} \label{tab:allpars}
\resizebox{\textwidth}{!}{
<<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=

@@

\begin{table}[ht]
\caption{Estimated latent variable means and covariances} \label{tab:lvpars}
\centering
<<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
require(xtable)

@@

\clearpage
\section{Summary statistics}
\begin{table}[ht]
\caption{Summary statistics} \label{tab:lvpars}
\centering
<<SUMSTATS, results='asis'>>=
require(dummies)

datacompl <- scores2plot[complete.cases(scores2plot[,decvars2]),] # complete cases
# expand dummies
datasum <- cbind(
  datacompl[,c("EXT","INT","cohortsex")], 
  dummy.data.frame(datacompl[,c(decvars2)], sep = "")
)

# statistics by group
sumtab <- data.frame(
  aggregate(datasum, by=list(datasum$cohortsex), FUN = function(x) c(mn = mean(x, na.rm=T)))
)

# drop unused factors and variables
sumtab <- sumtab[, substring(names(sumtab), nchar(names(sumtab))) != "0"]
sumtab <- sumtab[, ! names(sumtab) %in% c("EXT","INT","cohortsex","cohortBCS","cohortMCS")]
group <- sumtab$Group.1
sumtabt <- as.data.frame(t(sumtab[,-1]))
colnames(sumtabt) <- group
# rownames to var
sumtabt$var <- rownames(sumtabt)
# reorder columns
sumtabt <- sumtabt[c("var", "BCS.M", "MCS.M", "BCS.F", "MCS.F")]

# table
print.xtable(xtable(sumtabt,
                    align="L{0cm}L{3cm}C{1.2cm}C{2cm}C{2cm}C{2cm}",
                    digits = 3
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                   "& \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5}
                                    & BCS & MCS & BCS & MCS \\\\"
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabularx", width = "\\textwidth",
             floating = FALSE
)
